--- 
title: "Wildlife Presence Using Trail Cameras"
author: "Riley Robenstein"
date: "2025-12-08"
site: bookdown::bookdown_site
documentclass: book
csl: the-journal-of-wildlife-management.csl
bibliography: [references.bib, packages.bib]
url: https://rrobenstein.github.io/wildlife_presence_trail_cams/
description: "This project utilizes trail camera footage obtained from 2019-2023 in Idaho, Oregon, and Washington for Snapshot USA. The goal is to become familiar with assessing wildlife presence using trail camera data. Methods used here will help with assessing wildlife data (i.e., trail camera, acoustic recording, point count survey) for my research regarding deadwood quantification using LiDAR and wildlife presence and activity near deadwood structures."
link-citations: yes
github-repo: rrobenstein.github.io/wildlife_presence_trail_cams
---

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


```{r opts, echo = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```


# Remote sensing and wildlife

Use of remote sensing in wildlife research has become increasingly
prevalent, particularly with trail cameras. Trail cameras can provide
information about wildlife presence and absence when human observations
are unavailable or may disturb wildlife activity and presence. Continued
collection of data with trail cameras can provide spatial and temporal
trends of larger wildlife. The SNAPSHOT USA project contains several
years of camera trap data across the US available for public use
[@rooney2025b]. Data were collected using un-baited cameras and
identified through [Wildlife
Insights](https://app.wildlifeinsights.org/initiatives/2000156/Snapshot-USA).
The objective of this project is to become familiar with trail camera
data to assess wildlife presence and activity. This will ultimately help
in future data collection using trail cameras and acoustic recording
units deployed and point count surveys conducted in the University of
Idaho Experimental Forest to assess how deadwood structures may
influence wildlife presence and activity. This file assists in building
a database and cleaning and visualizing trail camera data.

## SNAPSHOT USA

SNAPSHOT USA is a national project where camera trap surveys are
conducted across all 50 US states annually from September-October to
estimate wildlife abundance and distributions [@rooney2025a]. This
project primarily focuses on mammalian species, with large birds also
included. SNAPSHOT USA's protocol provides a solid base for community
science, and various organizations across the country collaborate to
produce these spatial and temporal wildlife data. Some contributors have
provided data outside the standard collection period, so the 2019-2023
data span from August 1-December 29 [@rooney2025a]. Each collaborator is
required to deploy a camera array made up of a minimum of eight
un-baited cameras. The SNAPSHOT USA protocol requires at least 400
camera trap-nights between September and October per year for each
camera array [@rooney2025a].

## Study area

My focal area included a total of 15 trail camera sites in Idaho,
Washington, and Oregon (Figure 1).

![Figure 1. Camera array locations and Bailey's ecoregions of the US
from 2019-2023 [@rooney2024]. Darker points represent locations that fall within
Idaho, Washington and Oregon and size correlates to the number of years
surveyed. Colored polygons represent different
ecoregions.](documents/study_area.jpg)

## Data description

I subsetted these data to only include cameras deployed in Idaho,
Washington and Oregon. Data consisted of camera arrays, camera
locations, general habitat characteristics, deployment periods, image
sequences, and taxonomic information across various levels for species
identification [@rooney2025b]. Further information regarding camera
array locations and ecoregions were added for camera arrays
[@rooney2024].

## References

::: {#refs}
:::
